---
title: "Hand code on 1d SPDE"
author: "Elias T Krainski"
date: "September 2020"
output: html_document
bibliography: references.bib 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(INLA)
```

## Abstract

The aim of this tutorial covers three points. 
First is to define a SPDE model in INLA for 
one dimentional case. 
Second, is to perform Kriging using precision matrices. 
And, the third point is showing results 
considering two different set of basis functions chosen. 
Notice that this is just an illustration
and not how to proper do the analysis. 

## Introduction 

We consider the data analysied 
in the Gaussian Process example 
[here](https://people.bath.ac.uk/jjf23/brinla/gpreg.html). 

```{r data}
data(fossil, package='brinla')
fossil$sr <- (fossil$sr-0.7)*100
head(fossil, 3)
``` 

Selecting part of the data (shown in red)
```{r select}
set.seed(1)
isampl <- sort(sample(1:nrow(fossil), 30))
is.sel <- (1:nrow(fossil))%in%isampl
par(mar=c(2,3,0.5,0.5), mgp=c(2,0.5,0), las=1)
plot(fossil, pch=19, las=1, col=is.sel+1) 
fossil0 <- fossil
fossil <- fossil[isampl, ]
``` 

Define a grid over age, the set of knots for 
the basis functions. 

```{r knots}
range(fossil$age)
knots <- 2*(45:63) ## 90:125
length(knots)
```

The data model is of the following form 
\[y_i = \alpha + u_i + e_i \]
where $\alpha$ is the intercept, 
$u_i$ is the random field at the age 
for observation $i$, and $e_i$ is an 
independent error.

We re-write it in matrix formula as 
\[ \mathbf{y} = \mathbf{1} \alpha + \mathbf{A}\mathbf{u} + \mathbf{e} \]
where the vector $\mathbf{u}$ is the value of the 
random field at a set of knots placed over the range of age 
and $\mathbf{A}$ is a projector matrix. 
$\mathbf{A}$ projects the random field, 
modeled at the knots to the age of each observation. 

## The conditional mean (Kriging)

We consider $\alpha \sim N(0, 0^{-1})$, 
and $\mathbf{u} \sim N(\mathbf{0}, \mathbf{Q}_u^{-1})$, 
where $\mathbf{Q}_u$ is defined later. 

All we want to compute here is the conditional mean and 
variance of $\alpha$ and $\mathbf{u}$, conditional 
on the assumed prior distributions and the data $\mathbf{y}$. 
We can perform this following the Finn's material on GMRF 
[here](https://www.maths.ed.ac.uk/~flindgre/tmp/gmrf.pdf). 

We define $\mathbf{x} = \{\alpha, \mathbf{u}\}$, 
so that we have 
\[ 
\mathbf{Q}_x = \left[ 
\begin{array} {cc}
0 & 0 \\
0 & Q_u 
\end{array} \right] 
\]

Thus, we have
\[ \mathbf{x}|\mathbf{y},\theta \sim 
N(\mathbf{\mu}_{x|y,\theta}, 
\mathbf{Q}_{x|y,\theta}^{-1}) \]
where 
\[ \mathbf{Q}_{x|y,\theta} = 
\mathbf{Q}_x + 
\mathbf{A}^{T}\mathbf{Q}_e\mathbf{A}\] 
and 
\[ \mathbf{\mu}_{x|y,\theta} = 
\mathbf{Q}_{x|y,\theta}^{-1}
\mathbf{A}^{T}\mathbf{Q}_e\mathbf{y} 
\;.\] 

## One dimentional mesh 

First we use the defined knots in the 
`inla.mesh.1d()` to define the basis function setting. 

```{r}
m1 <- inla.mesh.1d(knots)
str(m1)
```

Them we can use this to define $\mathbf{A}$ 

```{r Ap}
Ap <- inla.mesh.projector(
    mesh=m1, loc=fossil$age)
str(Ap)

### or inla.spde.make.A(
##      mesh=m1, loc=$fossil$age)
A <- Ap$proj$A 

dim(fossil)
dim(A)
```

Showing $\mathbf{A}$ for some observations 
```{r Afew}
fossil$age[c(1,3,5)]
A[c(1,3,5), 1:10]
rowSums(A)
```

Visualizing the projector matrix (transposed)
```{r a1plot, fig.width=5, fig.height=3}
par(mar=c(0,0,0,0))
image(t(A), xlab='', ylab='', sub='') 
```

## The precision matrix for $\mathbf{u}$

We now have to define the precison of $\mathbf{u}$. 
For $\alpha=2$
\[\mathbf{Q}_2 = \tau^2 (\kappa^4 \mathbf{C} + 2\kappa^2\mathbf{G} + \mathbf{G}^{(2)})\]
where these matrices $\mathbf{C}$, $\mathbf{G}$ 
and $\mathbf{G}^{(2)}$ are defined in @lindgrenRL2011.

The $\kappa$ parameter is related to the practical range 
as $\kappa = \sqrt{8\nu}/\rho$ and 
the marginal variance as 
$\tau^2=\frac{\Gamma(\nu)}
{\Gamma(\alpha)(4\pi)^{d/2}\kappa^{2\nu}\sigma_u^2}$, 
@lindgrenR2015. 

The practical range and the marginal variance parameters 
are easier to undestand. 
After some "discussion" we can take 
$\sigma^2=0.5^2$ and $\rho=10$. 
We consider $\alpha=\nu+1/2=2$, giving $\nu=1.5$. 

```{r matrices}
alpha <- 2
d <- 1
nu <- alpha - d/2
rho <- 10
s2u <- 0.5^2
kappa <- sqrt(8*nu)/rho
tau <- sqrt(gamma(nu)/(2*sqrt(pi)*kappa^(2*nu)*s2u))
``` 

We can use the `inla.mesh.fem()` function to compute 
the matrices needed to build the precision matrix. 
```{r fem}
fe <- inla.mesh.fem(m1, order=2)
qx <- tau^2*(kappa^4 * fe$c0 + 2*kappa^2*fe$g1 + fe$g2)
image(qx)
```

Doing the same with the available functions in INLA
```{r qx1again}
spde1 <- inla.spde2.matern(
  mesh=m1, alpha=2) 
qx <- inla.spde2.precision(
  spde=spde1, theta=log(c(tau, kappa))) 
image(qx)
```

In addition, we have the noise variance error
```{r}
s2e <- 0.1^2
qe <- Diagonal(nrow(fossil), 1/s2e)
```

We stack the matrices to compute the 
conditional precision and mean

```{r mat1}
### XA = [1, A], in the material : BA
XA <- cbind(1, A)

### Qx = [ block diagonal ]
Qx <- cbind(0, rbind(0, qx))
dim(qx)
dim(Qx)
```

The precision is 
```{r prec1}
Qx.y <- Qx + crossprod(XA, qe)%*%XA
dim(Qx.y)
```

The conditional mean can be computed using 
```{r mu1}
mx.y <- drop(inla.qsolve(
  Qx.y, crossprod(XA, qe)%*%fossil$sr))
```

The prediction can be computed, and visualized, with
```{r ypred}
plot(fossil)
lines(fossil$age, drop(XA %*% mx.y))
```

The conditional variance standard error 
```{r var1}
Vx <- inla.qinv(Qx.y)
v.x <- diag(Vx)
s.x <- sqrt(v.x)
```

Adding the error in the plot
```{r viz1}
plot(fossil, las=1)
lines(knots, mx.y[1]+mx.y[-1])
lines(knots, mx.y[1]+mx.y[-1] - 1.96*s.x[-1], lty=2)
lines(knots, mx.y[1]+mx.y[-1] + 1.96*s.x[-1], lty=2)
```

## Second order basis functions 

The piecewise linear basis functions 
is just one way to do do this 
and there are several options to this. 
A comparison among quite a few ones 
was performed in @bolinL2011. 
In general, one has to balance between the properties 
of the chosen basis functions and how sparse 
$\mathbf{Q}_{x} + \mathbf{A}^T\mathbf{Q}_e\mathbf{A}$ is.

Next we do the same computations using 
second order B-splines. 

```{r A2}
m2 <- inla.mesh.1d(knots, degree=2)
str(m2)
A2 <- inla.spde.make.A(
    mesh=m2, loc=fossil$age)
fossil$age[c(1,3,5)] 
A2[c(1,3,5), 1:7]

table(rowSums(A2))
table(rowSums(A2>0))
```

We can visualize how the 2nd degree basis 
functions are with 
```{r viz2nd}
loc0 <- seq(90, 125, 0.1)
aplot <- inla.spde.make.A(
    mesh=m2, loc=loc0)

plot(loc0, aplot[,1], type='n')
for (j in 1:ncol(aplot))
    lines(loc0, aplot[,j])
```

Now we compute the conditional precision and mean again. 
We will use the `inla.spde2.matern()` and 
`inla.spde2.precision()` functions to help 
building $\mathbf{Q}_x$ at this time
```{r q2}
spde2 <- inla.spde2.matern(
  mesh=m2, alpha=2)
q2x <- inla.spde2.precision(
  spde=spde2, theta=log(c(tau, kappa)))
### Qx
Q2x <- cbind(0, rbind(0, q2x))

XA2 <- cbind(1, A2)
Q2x.y <- Q2x + crossprod(XA2, qe)%*%XA2

m2x.y <- drop(inla.qsolve(
  Q2x.y, crossprod(XA2, qe)%*%fossil$sr))

## variance
V2x <- inla.qinv(Q2x.y)
v.x2 <- diag(V2x)
s.x2 <- sqrt(v.x2)
```

We now visualize booth results 
```{r viz2}
plot(fossil0, pch=19, las=1, col=is.sel+1) 

lines(knots, mx.y[1]+mx.y[-1])
lines(knots, mx.y[1]+mx.y[-1] - 1.96*s.x[-1], lty=2)
lines(knots, mx.y[1]+mx.y[-1] + 1.96*s.x[-1], lty=2)

lines(m2$mid, m2x.y[1]+m2x.y[-1], col=2)
lines(m2$mid, m2x.y[1]+m2x.y[-1] - 1.96*s.x2[-1], lty=2, col=2)
lines(m2$mid, m2x.y[1]+m2x.y[-1] + 1.96*s.x2[-1], lty=2, col=2)
```


# References
